When to trust phylogeographic inferences?

### 2024-05-07 manuscript/phylogeo.tex
manuscript edits

\maketitle
As organisms replicate, they also disperse in space, potentially exploring new habitats.
The reconstruction of ancestral locations and past migrations from sampling locations of extant individuals, typically in combination with genome sequence information to infer the phylogeny, is known as phylogeography.
Phylogeographic methods are implemented in popular evolutionary analysis software such as BEAST \citep{pybus_unifying_2012}.
The underlying model of dispersal is typically diffusive, that is the probability of sampling a descendant at position $\rvec_c$ after time $t$ when the parent was at position $\rvec_p$ is given by
\begin{equation}
    P(\rvec_c| \rvec_p, t) = \frac{1}{4\pi D t}e^{\frac{|r_c - r_p|^2}{4Dt}}
\end{equation}
where $D$ is a diffusion constant with dimensions $\mathrm{length}^2/\mathrm{time}$, and we have assumed a two-dimensional space.
Each branch in the phylogeny is decorated with such a diffusive propagator.
Such a diffusive process is the simplest and most natural choice in absence of directed motion or long range migration.
It is also mathematically convenient as unknown ancestral locations can be integrated in closed form and the marginal distribution of each position can be calculated exactly for a fixed tree along with the likelihood. 
The total likelihood of the configuration is then assumed to be the product of the spatial likelihood and the phylogenetic likelihood, i.e. tree topology and branching rates of the tree are assumed to be independent of the spatial process.
However, these two interact in important ways, for example when a population expands into an uninhabited fertile habitat or a disease spreads in a susceptible population.

In a typically Bayesian phylogeographic inference problem as discussed in \citet{pybus_unifying_2012}, the input data are typically viral genome sequences as well as their sampling times and sampling locations.
Software like BEAST will then sample the posterior distribution of the phylogenetic tree topology, the sampling times and locations of ancestral nodes in the tree, and model parameters like the diffusion constant $D$ and the evolutionary rate $\mu$.
Results can be estimates of the ancestral location of a particular groups of samples, or summary statistics of the dispersal process.
Popular summaries include empirical estimates of the diffusion constant  \citep{pybus_unifying_2012} and a so-called \emph{dispersal velocity} \citep{dellicour_using_2017}. These are calculated from the estimated displacements along branch $i$ with parent $p_i$ given by $d_i = |\rvec_{p_i} - \rvec_{i}|$ and the elapsed time $\Delta t_i$ as
\begin{equation}
    D_w = \frac{\sum_{i\in B}d_i^2}{4\sum_{i\in B} \Delta t_i} \quad \mathrm{and}  \quad v_w = \frac{\sum_{i\in B} d_i}{\sum_{i\in B} \Delta t_i}
\end{equation}
where $B$ is the set of all branches of the tree.
Effectively $D_w$ divides the sum of all observed squared displacements by the total time elapsed on the tree to obtain and estimate of $D$.
It is known as the `weighted' estimate of the diffusion constant.
The weighted dispersal velocity estimate compares the sum of absolute values of observed displacements to the total time, which has dimensions of a velocity.
Alternative formulations use the mean of fractions $D_b = |B|^{-1} \sum_{i\in B}d_i^2/4t_i$ or $v_b = |B|^{-1} \sum_{i\in B}d_i/t_i$ instead of the ratio of sums. \
We refer to these here as `by branch' estimates. 
The latter tend to be dominated by short branches and thus noisier.

@@

Fig.~\ref{fig:D_and_v} shows empirical estimates of diffusion constants and velocity as defined above from simulated data using freely diffusing locations along Kingman trees.
The diffusion constant $D$ can be estimated robustly from these data and are compatible with the diffusion constant used to simulate the data $D=1$.
Values of $v_w$ and $v_b$, however, depend strongly on the sample size. 
That the velocity estimates are ill-defined is not surprising: By changing the sample size, the average length of branches changes. Diffusion along each branch results in a displacement of $d_i \sim \sqrt{4Dt_i}$ such that a quantity that depends on $\Delta d_i / t_i \sim 1/t_i^{1/2}$ will change with the time $t_i$ that elapsed along the branch.
With more samples in the tree, the average branch length decreases and the velocity estimates increase. 
This expected scaling $\sim \sqrt{n}$ is indicated by the dashed line in Fig.~\ref{fig:D_and_v}.
For Yule trees, the average branch length is independent of the number of samples and the velocity estimates are more stable. 
But estimates for $v_w$ and $v_b$ are not consistent with each other and their numerical values have no interpretation within the model.

In either case, estimates of $v_w$ or $v_b$ are not meaningful. 
The underlying model is diffusive and has no parameters that individually or combined would have dimensions of a velocity. 
There is no ground truth to compare the estimates to and $v_w$ and $v_b$ are just descriptive summaries of spatial spread with unclear interpretation.

@@

\section*{How does density regulation affect estimates phylogeographic estimates?}
Phylogeographic inference typically assumes that the spatial diffusion process is independent of the branching pattern of the tree and ignores coupling between the two. 
Such a coupling could, for example, arise when growth rates depend on local population density, and it has been known for a long time that ignoring such coupling leads to very unrealistic population densities \citep{felsenstein_pain_1975}.
If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D<L^2/T_c$, where $T_c$ is the coalescent times scale of the population (see Fig.~\ref{fig:heterogeneity}).
In a well-mixed neutral model, $T_c$ is equal to the population size $N$.

@@

\begin{figure*}[tb]
    \includegraphics*[width=\textwidth]{figures/heterogeneity_free_diffusion.pdf}
    \caption[short]{\label{fig:heterogeneity}{\bf Population density variation in free diffusion models.}
    When spatial motion is independent of population density and coalesce is sufficiently rapid, the realized population density fluctuates strongly across the habitat. The left panel shows the population density for the case $D = 0.1 L^2/N$ in linear bins of size $L/20$. The right panels shows the standard deviation of the population density for different bin sizes as a function of the diffusion coefficient. For $D>L^2/N$, the heterogeneity approaches the expected value in the well mixed limit. }
\end{figure*}

@@

As soon as the interaction radius $r$ is much smaller than the habitat size, this density regulation leads to a more homogeneous population density, see Fig.~\ref{fig:density_reg}.
Diffusion constants estimated from the population remain accurate with moderate overestimation of small diffusion constants.
But with decreasing diffusion, the population fragments into smaller than smaller subpopulations that don't interact or coalesce.
This manifests in an increasing time to the MRCA, see Fig.~\ref{fig:density_reg}.
Spatial location and the coalescence process are thus strongly coupled, though this coupling manifests itself in population structure rather than estimates of the diffusion coefficients.
In such a spatially structured population, estimates of historical populations sizes (e.g.~skyline estimates) will be strongly affected.

@@

To explore such the result of such changing habitats on phylogeographic inferences, we simulated a habitat with a Gaussian carrying capacity that moves at a constant velocity $v$ along the $x$-axis (Fig.~\ref{fig:traveling_wave}). 
When varying the microscopic dispersal parameter $D$ and the velocity of the carrying capacity $v$, I observe diverse behavior with often dramatically overestimated $\hat{D}$. These behaviors are explained comparing the speed $v_{FKPP}$ at which the population can invade new territory with the external speed $v$, the ratio of which is used to plot the simulation data in Fig.~\ref{fig:traveling_wave}. 
For $v_{FKPP} \ll v$, the population cannot keep up with the moving carrying capacity and is saved from extinction only because selection is soft.
In this regime, the estimated velocity of lineages is below $v$ and both the estimated $v$ and $D$ increase with $v_{FKPP}$. 
Once $v\sim v_{FKPP}$, the estimated lineage velocity coincides with $v_{FKPP}$, while the estimated $D$ peaks at values orders of magnitude above the true value. 
For larger $v_{FKPP}$, the estimated velocity is stable for some time, before increasing again with $v_{FKKP}$ is a sampling dependent manner. 
At larger $v_{FKPP}$, $D$ is eventually underestimates.

>>>>

\maketitle
As organisms replicate, they also disperse in space, potentially exploring new habitats.
The reconstruction of ancestral locations and past migrations from sampling locations of extant individuals, typically in combination with genome sequence information to infer the phylogeny, is known as phylogeography.
Phylogeographic methods are implemented in popular evolutionary analysis software such as BEAST \citep{pybus_unifying_2012}.
The underlying model of dispersal is typically assumed to be diffusive, that is the probability of sampling a descendant at position $\rvec_c$ after time $t$ when the parent was at position $\rvec_p$ is given by
\begin{equation}
    P(\rvec_c| \rvec_p, t) = \frac{1}{4\pi D t}e^{\frac{|r_c - r_p|^2}{4Dt}}
\end{equation}
where $D$ is a diffusion constant with dimensions $\mathrm{length}^2/\mathrm{time}$, and we have assumed a two-dimensional space.
Such a diffusive process is the simplest and most natural choice in absence of directed motion or long range migration.
It is also mathematically convenient as unknown ancestral locations can be integrated in closed form and the marginal distribution of each position can be calculated exactly for a fixed tree along with the likelihood.
The total likelihood of the configuration is then assumed to be the product of the spatial likelihood and the phylogenetic likelihood, i.e. tree topology and branching rates of the tree are assumed to be independent of the spatial process.

However, population dynamics is often tightly coupled to spatial location, for example when a population expands into an uninhabited fertile habitat or a disease spreads in a susceptible population, and the independence assumption of spatial and phylogenetic processes is not justified.
The extent to which violations of this assumption affect phylogeographic inferences is not well understood.
To accommodate variation not captured by the simple diffusion model, ``relaxed random walk'' models allow the diffusion constant $D$ to vary across the tree (the diffusion constant is integrated over a prior distribution) \citep{dellicour_relax_2021}.

In a typically Bayesian phylogeographic inference problem as discussed in \citet{pybus_unifying_2012}, the input data are typically viral genome sequences as well as their sampling times and sampling locations.
Software like BEAST will then sample the posterior distribution of the phylogenetic tree topology, the sampling times and locations of ancestral nodes in the tree, and model parameters like the diffusion constant $D$ and the evolutionary rate $\mu$.
Results can be estimates of the ancestral location of a particular groups of samples, or summary statistics of the dispersal process.
Popular summaries include empirical estimates of the diffusion constant  \citep{pybus_unifying_2012} and a so-called \emph{dispersal velocity} \citep{dellicour_using_2017}. These are calculated from the estimated displacements along branch $i$ with parent $p_i$ given by $d_i = |\rvec_{p_i} - \rvec_{i}|$ and the elapsed time $\Delta t_i$ as
\begin{equation}
    \label{eq:disperal_parameters}
    D_w = \frac{\sum_{i\in B}d_i^2}{4\sum_{i\in B} \Delta t_i} \quad \mathrm{and}  \quad v_w = \frac{\sum_{i\in B} d_i}{\sum_{i\in B} \Delta t_i}
\end{equation}
where $B$ is the set of all branches of the tree.
Effectively $D_w$ divides the sum of all observed squared displacements by the total time elapsed on the tree to obtain and estimate of $D$.
It is known as the `weighted' estimate of the diffusion constant.
The weighted dispersal velocity estimate compares the sum of absolute values of observed displacements to the total time, which has dimensions of a velocity.
Alternative formulations use the mean of fractions $D_b = |B|^{-1} \sum_{i\in B}d_i^2/4t_i$ or $v_b = |B|^{-1} \sum_{i\in B}d_i/t_i$ instead of the ratio of sums. \
We refer to these here as `by branch' estimates.
The latter tend to be dominated by short branches and thus noisier.

@@

Fig.~\ref{fig:D_and_v} shows empirical estimates of diffusion constants and velocity as defined in Eq.~\ref{eq:disperal_parameters} from simulated data using freely diffusing locations along Kingman trees.
The diffusion constant $D$ can be estimated robustly from these data and the estimates are compatible with the diffusion constant used to simulate the data $D=1$.
Values of $v_w$ and $v_b$, however, depend strongly on the sample size.
That the velocity estimates are ill-defined is not surprising: By changing the sample size, the average length of branches changes. Diffusion along each branch results in a displacement of $d_i \sim \sqrt{4Dt_i}$ such that a quantity that depends on $\Delta d_i / t_i \sim 1/t_i^{1/2}$ will change with the time $t_i$ that elapsed along the branch.
With more samples in the tree, the average branch length decreases, and the velocity estimate increases.
This expected scaling $\sim \sqrt{n}$ is indicated by the dashed line in Fig.~\ref{fig:D_and_v}.
For Yule trees, the average branch length is independent of the number of samples and the velocity estimates are more stable.
But estimates for $v_w$ and $v_b$ are not consistent with each other.

In either case, estimates of $v_w$ or $v_b$ are not meaningful.
The underlying model is diffusive and has no parameters that individually or combined would have dimensions of a velocity.
There is no ground truth to compare the estimates to and $v_w$ and $v_b$ are just descriptive summaries of spatial spread that can not be compared between datasets.

@@

\section*{How does density regulation affect estimates phylogeographic estimates?}
Phylogeographic inference typically assumes that the spatial diffusion process is independent of the branching pattern of the tree and ignores coupling between the two.
Such a coupling, however, is expected since growth rates naturally depend on local population density when organisms compete for resources.
It has been known for a long time that ignoring such coupling leads to unrealistically heterogeneous population densities \citep{felsenstein_pain_1975}.
If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D<L^2/T_c$, where $T_c$ is the coalescent timescale of the population (see Fig.~\ref{fig:heterogeneity}).
In a well-mixed neutral model, $T_c$ is equal to the population size $N$.

@@

\begin{figure*}[tb]
    \includegraphics*[width=\textwidth]{figures/heterogeneity_free_diffusion.pdf}
    \caption[short]{\label{fig:heterogeneity}{\bf Population density variation in free diffusion models.}
    When spatial motion is independent of population density and coalesce is sufficiently rapid, the realized population density fluctuates strongly across the habitat. The left panel shows the population density for the case $D = 0.1 L^2/N$ in linear bins of size $L/20$. The right panel shows the standard deviation of the population density for different bin sizes as a function of the diffusion coefficient. For $D>L^2/N$, the heterogeneity approaches the expected value in the well mixed limit. }
\end{figure*}

@@

As soon as the interaction radius $r$ is much smaller than the habitat size, this density regulation leads to a more homogeneous population density, see Fig.~\ref{fig:density_reg}.
Diffusion constants estimated from the population using Eq.~\ref{eq:dispersal} remain accurate with moderate overestimation of small diffusion constants.
If instead of periodic boundary conditions reflective boundary conditions are used, diffusion constants are underestimated once $\sqrt{2DT_c}$ becomes comparable to the habitat size.
This underestimation will then result in over-confident inference of ancestral locations, in particular for recent ancestral nodes.

But with decreasing diffusion, the population fragments into smaller than smaller subpopulations that don't interact or coalesce.
This manifests in an increasing time to the MRCA, see Fig.~\ref{fig:density_reg}.
Spatial location and the coalescence process are thus strongly coupled, though this coupling manifests itself in population structure rather than estimates of the diffusion coefficients.
In such a spatially structured population, estimates of historical populations sizes (e.g.~skyline estimates) will be strongly affected.

@@

To explore such the result of such changing habitats on phylogeographic inferences, we simulated a habitat with a Gaussian carrying capacity that moves at a constant velocity $v$ along the $x$-axis (Fig.~\ref{fig:traveling_wave}) with periodic boundary conditions.
When varying the microscopic dispersal parameter $D$ and the velocity $v$ of the carrying capacity profile, I observe diverse behavior with often dramatically overestimated $\hat{D}$. These behaviors are explained comparing the speed $v_{FKPP}$ at which the population can invade new territory with the external speed $v$, the ratio of which is used to plot the simulation data in Fig.~\ref{fig:traveling_wave}.
For $v_{FKPP} \ll v$, the population cannot keep up with the moving carrying capacity and is saved from extinction only because selection is soft.
In this regime, the estimated velocity of lineages is below $v$ and both the estimated $v$ and $D$ increase with $v_{FKPP}$.
Once $v\sim v_{FKPP}$, the estimated lineage velocity coincides with $v_{FKPP}$, while the estimated $D$ peaks at values orders of magnitude above the true value.
For larger $v_{FKPP}$, the estimated velocity is stable for some time, before increasing again with $v_{FKKP}$ is a sampling dependent manner.
At larger $v_{FKPP}$, $D$ is eventually underestimates.
